---
title: "Part 3: Generalized Linear Models"
author: "Charles Julien, Chike Odenigbo, Atul Sharma, Gabriel Jobert"
date: "11/17/2023"
geometry: "left=2cm,right=2cm,top=1cm,bottom=1.2cm"
output:
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, warning=FALSE, include = FALSE, echo=FALSE}
# Import (The fewer the better)

library(tidyverse)
library(ggplot2)
library(dplyr)
library(leaflet)
library(gridExtra)
library(broom)
library(stats)
library(car)
library(janitor)
library(tibble)
library(MASS)
library(AER)
```

```{r load_data, include=FALSE}
# Data loading

# Set working directory to file location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
df_raw = read.csv("./../src/bixi10.csv")
df_station = read.csv("./../src/2021_stations.csv")
df_main = left_join(df_raw,df_station,by = c("station"="pk"))
head(df_main)

```

```{r echo=FALSE}
# Data preparation and feature engineering


# Note: Correlation between rev and dur is very high 0.994467

df_main <- df_main %>% mutate(
              holiday = factor(holiday),
              mem = factor(mem),
              mm = factor(mm),
              station = factor(station),
              
              wday_ordered = factor(wday, levels = c("Monday", "Tuesday", "Wednesday",  "Thursday", "Friday", "Saturday", "Sunday"),ordered = TRUE),
              season = ifelse((mm %in% c(12,1,2)), 'Winter',
                       ifelse((mm %in% c(3,4,5)), 'Spring',
                       ifelse((mm %in% c(6,7,8)), 'Summer',
                       ifelse((mm %in% c(9,10,11)), 'Fall','No Season')))),
              
              rain_ind = factor(ifelse((rain != 0), 'Rain', 'NoRain')),
              
              # Compared to Parc Lafontaine 
              North_South = factor(ifelse(latitude>45.5271, 'North', 'South')),
              West_East = factor(ifelse(longitude>-73.5705, 'East', 'West')),
              
              # Beginning of month versus end of month
              PartOfMonth = factor(ifelse(dd>15, 'EOM', 'BOM')),
              
              Metro_ind = factor(ifelse(grepl("Métro", name),1,0)),
              
              lntot = log(n_tot),
              
              rev_per_min = rev/dur,
              rev_per_trip = rev/n_tot,
             
              percent_AM = n_AM/n_tot,
              percent_PM = n_PM/n_tot,
              
              percent_AM_PM_delta = percent_AM - percent_PM,
              n_AM_PM_delta = n_AM - n_PM,
              
              am_pm_ind = ifelse(n_AM>=n_PM, 'AM','PM'),
              
              wknd_ind = factor(ifelse((wday %in% c("Sunday","Saturday")), 'Weekend'
                        ,'Weekday')),
                
              long_wknd_ind = factor(ifelse((wday %in% c("Sunday","Saturday")), 'Weekend'
                        ,ifelse((wday %in% c("Friday","Monday") & holiday=='1'),'Long Weekend','Weekday'))),
        
              year = 2021 ,
              
              date = as.Date(paste(year,mm,dd,sep="/")),
              
              week_num = as.numeric(strftime(date, format = "%V")),
 

# create a new variable indicating whether the average daily trip duration exceeds 15 minutes, and explore models for this new variable.

              avg_15_ind = factor(ifelse(avg>15,1,0))

)


```

# Instructions

Part 3: Generalized linear models (due November 10 before 11:55 PM)

• Explore various generalized linear models for the response variables
of interest, specifically, for the number of rentals (total, AM, and
PM). In addition, create a new variable indicating whether the average
daily trip duration exceeds 15 minutes, and explore models for this new
variable.

• Be sure that your analyses allow you to answer well formulated
business / research questions that you wish to address. The goal is to
use generalized linear models to provide interesting and relevant
insights from the data.

• Comment on findings and discuss the main takeaways from this analysis
from a business perspective. Be sure to provide relevant model outputs
that support your discussion.

• Discuss any shortcomings or limitations of the analyses carried out.

# Introduction

# Business/Research questions

The target variables is number of rentals (total, AM, and PM)

# Pre-processing

## Imputation (might have to remove)

Revenue for members is missing since they do not pay a usage fee, but
rather a fixed cost.

```{r}
imputation_model <- lm(rev ~ dur + avg + n_tot, data = df_main)
df_main$rev_pred = predict(imputation_model, df_main)
```

To impute revenue for members, we make the assumption that they would
bring in as much revenue as non-members for the same usage.Thus we
consider the same formula of revenue used for non-members. This unknown
deterministic function is most likely a linear combination of usage
variables like `dur`, `avg` and `n_tot`. We try to approximate this
function and use it to impute members revenue. The imputation model has
an r-squared of 1 on non-members data.

# Research Question 1: How does membership and holidays affect likelihood of longer trips (average duration exceeds 15 minutes) ?

**Objective of Analysis:** Understand member behavior in terms of rental
duration to tailor membership benefits and pricing strategies.

## Variables Selection

To evaluate how the membership and holidays affect likelihood of longer
trips, we need to create a model that accounts for the membership and
holidays variable and to use the new variable created to know if the
trip is above 15 minutes as the interest variable. The goal would be to
quantify the relationship between these factor and the "longer trips"
variable.

## Model

Here we use a logistic regression model, which is a type of GLM suitable
for binary outcomes, with a binomial distribution and a logit link
function.

```{r model1, echo=FALSE}
glm_membership <- glm(avg_15_ind ~ mem + holiday + long_wknd_ind, family = binomial, data = df_main)
summary(glm_membership)
```

## Interpretation

-   **Intercept (`-0.72009`)**: This is the log odds of the average trip
    duration exceeding 15 minutes for a non-member (because `mem` is 0
    for non-members), on a day that is neither a holiday (`holiday` is
    0)  nor part of a long weekend (`long_wknd_ind` is 0). The negative
        value of the intercept suggests that, for this reference group,
        the probability of having a trip duration exceeding 15 minutes
        is less than 50%. To convert the log odds to probability, you
        would use the logistic function
        :$P(\text{avg duration} > 15 | \text{non-member, non-holiday, non-long weekend}) = \frac{1}{1 + e^{0.72009}}$We
        can calculate the exact probability:
        $$P = \frac{1}{1 + e^{-(-0.72009)}} \approx \frac{1}{1 + e^{0.72009}} \approx \frac{1}{1 + 2.054} \approx \frac{1}{3.054} \approx 0.3275$$
        So, the estimated probability of a non-member taking a trip
        longer than 15 minutes on a regular day (not a holiday and not a
        long weekend) is approximately 32.75%.
-   **Membership (`mem1`, `-0.88294`)**: The negative coefficient for
    **`mem1`** indicates that BIXI members are less likely to have trips
    that exceed 15 minutes on average, compared to non-members. The odds
    ratio can be calculated as $e^{-0.88294} \approx 0.414$, which means
    the odds of members taking longer trips are about 41.4% of the odds
    for non-members. Members are therefore less likely to take longer
    trips, which could indicate that members are using the service for
    shorter, more regular commutes.
-   **Holiday (`holiday1`, `0.98469`)**: The positive coefficient for
    **`holiday1`** indicates that on holidays, the odds of trips
    exceeding 15 minutes are higher. The odds ratio is
    $e^{0.98469} \approx 2.677$, suggesting that the likelihood of
    longer trips is about 167.7% higher on holidays than on
    non-holidays.
-   **Long Weekend Weekday (`long_wknd_indWeekday`, `0.58520`)**: The
    positive coefficient here suggests that on weekdays that are part of
    a long weekend, the likelihood of longer trips increases. The odds
    ratio is $e^{0.58520} \approx 1.796$, indicating that the odds are
    about 79.6% higher on these days compared to regular weekdays.
-   **Long Weekend Weekend (`long_wknd_indWeekend`, `1.14434`)**:
    Thisvcoefficient is even more substantial, with an odds ratio of
    $e^{1.14434} \approx 3.140$, suggesting that the odds of longer
    trips on weekends that are part of a long weekend are more than
    three times higher compared to a non-long weekend day.
-   **Statistical Significance**: In this model, all the p-values are
    under the 5% level of significance, indicating that the relationship
    between these variables and the likelihood of taking a longer trip
    are statistically significant.
-   **Model Fit**: The AIC of the model is 12867, which can be used for
    model comparison purposes. The lower the AIC, the better the model
    fits the data while penalizing for complexity.

## Business implications (can change the sub categories)

-   Since members are less likely to take longer trips, membership
    benefits and pricing could be adjusted to encourage more extended
    use, or to better cater to the frequent, shorter trips that members
    seem to prefer.

-   The increase in longer trips during holidays and long weekends
    indicates potential opportunities for targeted marketing and
    promotions to encourage bike usage during these periods.

-   The significant increase in the likelihood of longer trips during
    long weekends, especially on the weekend days, suggests that there
    might be a need for increased bike availability and maintenance
    during these times to accommodate the higher demand for leisurely
    rides.

# Research Question 2: How do the seasonal variation in weather affect the total number of trip ?

**Objective of Analysis:** Evaluate how seasonal weather patterns
influence rental numbers to inform seasonal staffing and maintenance
schedules.

## Variables Selection

To evaluate how seasonal weather patterns influence rental numbers, we
need to create a model that accounts for the various factors that can
vary with seasons, such as temperature, rainfall and specific time of
year (season). The goal would be to quantify the relationship between
these factors and the number of rentals, which can then inform decisions
on staffing and maintenance schedules.

## Model

Rental numbers are count data, so a Poisson or negative binomial GLM
would both be suitable. Here, we first fit a Poisson regression model
and check for over-dispersion. Since the value is significantly greater
than 1, we then fit a Negative Binomial model that will be more
appropriate for this particular case.

```{r model_2, echo=FALSE}
# Poisson model
glm_seasonal_weather <- glm(n_tot ~ temp + rain + season, family = poisson, data = df_main)
dispersiontest(glm_seasonal_weather)

# Negative Binomial model
glm_seasonal_weather_2 <- glm.nb(n_tot ~ temp + rain + season, data = df_main)
summary(glm_seasonal_weather_2)
```

## Interpretation

-   **Theta**: The estimate for theta ($\theta = 0.9902$) suggests that
    the variance is slightly less than the mean, which justifies the use
    of the Negative Binomial model over the Poisson model.

-   **Model Fit**: The AIC of the model is 79919, which can be used for
    model comparison purposes. The lower the AIC, the better the model
    fits the data while penalizing for complexity.

-   **Intercept**: The intercept ($\beta = 2.581791$) represents the log
    count of the total number of trips when all other variables are zero
    (which is not possible for temperature or season, but serves as a
    reference point). The IRR for the intercept cannot be interpreted in
    the same way because it relates to the situation where all predictor
    variables are zero, which may not be meaningful for variables like
    temperature.

-   **Temperature (`temp`)**: The incidence rate ratio (IRR) for
    temperature is $e^{0.031157}$, which is approximately 1.032. This
    means that for each one-degree Celsius increase in temperature, the
    expected number of total trips increases by a factor of 1.032, or
    3.2%.

-   **Rainfall (`rain`)**: The IRR for rainfall is $e^{-0.015916}$,
    which is approximately 0.984. This indicates that for each
    additional millimeter of rainfall, the expected number of trips
    decreases by a factor of 0.984, or 1.6%. So, if rainfall increases
    by 1 mm, the model predicts a 1.6% decrease in the number of trips.

-   **Season (`seasonSpring`, `seasonSummer`)**:

    -   For `seasonSpring`: The IRR is $e^{-0.265988}$, approximately
        0.767. This suggests that, all else being equal, the expected
        number of trips in spring is 76.7% of the number in the baseline
        season (`seasonFall`), which is a 23.3% decrease.

    -   For `seasonSummer`: The IRR is $e^{-0.113913}$, approximately
        0.892. This means that in summer, the expected number of trips
        is 89.2% of the number in the baseline season (`seasonFall`), a
        10.8% decrease.

## Business implications

-   The bike-sharing service is likely to see increased demand on
    warmer, drier days. This can guide the allocation of bikes across
    stations and the scheduling of staff for redistribution and customer
    service.

-   During rainy days, demand is expected to drop, which could be a good
    time for scheduling maintenance work.

-   The unexpected decrease in trips during spring and summer compared
    to the baseline season suggests that additional factors might need
    to be considered, or specific marketing strategies might be
    implemented to boost ridership during these seasons.

# Research Question 3: What variables impacts the porportion of trips in the morning versus in the evening and in what way ?

**Objective of Analysis:** The goal of this analysis is to understand
what variables influence the repartition of the trips throughout the
day. Knowing this would help to better forecast the demand for bikes
across the bixi system.

Before starting the analysis, it is important to know that our datasets
has 57.76% of its trips in the afternoon.

```{r}
sum(df_main$n_PM)/sum(df_main$n_tot)
```

## Variables Selection

Some variables that would be interesting to investigate are the
following:

`mem` : Membership indicator

`wknd_ind` : Indicator of weekend

`season` : Categorical variable with autumn, summer and fall

`temp` : Temperature in degrees celcius

`rain` : Precipitation in mm

`North_South` : Indicator of cardinality compared to parc lafontaine

`West_East`: Indicator of cardinality compared to parc lafontaine

`Metro_ind` : Indicator of metro station nearby

**Correlation:**

The correlation between the variables selected above was already tested
via multicollinearity tests like VIF in previous analysis and did not
show source of concern.

**Interest variable:**

Our interest variable for this question is the proportion of trips in
the afternoon compared to the total number of trips. Since `n_PM` is a
count we expect a poisson like distribution. To obtain a rate we will
use the variable `n_tot` as an offset.

```{r interest variable}
hist(df_main$n_PM, breaks = 30)
```

```{r dispersion}
mean(df_main$n_PM)
var(df_main$n_PM)
```

We observe some big disparities between the mean and variance of the
variable `n_PM` which could lead to some overdispersion in our model.
Some formal test will be explored in the model part.

## Model

```{r model_3, echo=FALSE}
mod.poi <- glm(n_PM ~ mem + wknd_ind +season + temp + rain + North_South  + West_East + Metro_ind + offset(log(n_tot)) , data = df_main, family = poisson(link = "log"))

summary(mod.poi)

```

```{r}
# Deviance based
mod.poi$deviance/mod.poi$df.residual

# Based on Pearon X2 statistic
sum(residuals(mod.poi, type = "pearson")^2)/mod.poi$df.residual
```

Once the covariates and offset are taken into consideration, we seem to
be more in a case of underdispersion since theta is smaller than 1. For
this reason, we will explore a quasipoisson distribution in order to
increase flexibility and allow the mean to be different from the
variance.

```{r}
mod.quasi <- glm(n_PM ~ mem + wknd_ind +season + temp + rain + North_South  + West_East + Metro_ind + offset(log(n_tot)) , data = df_main, family = quasipoisson)
summary(mod.quasi)
```

```{r}
Anova(mod.quasi, type = 3)
```

## Interpretation

**Overall Model** We observe that the deviation parameter of the model
is estimated to be 0.68 which means that the means given the covariates
is bigger than its variance. According to the Anova command, all
variables included in the model are significant, once adjusted for all
the covariates, except for `metro_ind`

**Intercept** : -0.605 can be interpreted as the log average rate of
trip in the afternoon when all covariates have values of zero.

In other words, this is the average rate of trip in the afternoon for
non-member, during the weekday, in fall, when temperature is zero
degrees celcius, no precipitation, with rental location north east of
parc lafontaine and at an acces point that as no metro station nearby
which is exp(-0.605) = 54.6%.

**Membership :** 0.039 is a positive coefficient hence membership
increase the rate of trip in the afternoon, on average, when all else is
being held constant by a factor of exp(0.039) = 1.039 which means an
increase of about 4% .

**Weekend:** 0.022 is a positive coefficient hence weekend increase the
rate of trip in the afternoon, on average, when all else is being held
constant by a factor of exp(0.022) = 1.022 which means an increase of
about 2% .

**Season:** Spring's coefficient is 0.181 and summer's is 0.005 meaning
that they both have an increase in rate of trip in the afternoon on
average compared to fall when all else is being held constant. This
increase is of a factor of exp(0.181) = 1.198 and exp(0.005) = 1.005
respectively for both season.

**Temperature:** -0.002 which means that a one degree celcius increase
in temperature results in a decrease in rate of afternoon trips on
average when all else is being held constant. This decrease is of a
factor of exp(-0.002) = 0.998.

**Precipitation :**- rain coefficient is also -0.002 hence its
interpretation is the same as for temperature except that the decrease
is for each additional milliliters of rain.

**Cardinality North-South:** 0.041 is a positive coefficient hence
departure from bixi station south of parc lafontaine have a higher rate
of trip in the afternoon, on average, when all else is being held
constant by a factor of exp(0.041) = 1.041 which means an increase of
about 4% compared to northern departure.

**Cardinality West-East:** -0.017 is a positive coefficient hence
departure from bixi station west of parc lafontaine have a lower rate of
trip in the afternoon, on average, when all else is being held constant
by a factor of exp(-0.017) = 0.983 which means a decrease of about 2%
compared to eastern departure.

**Metro station nearby:** coefficient is 0.009 which although not being
significatively differrent from zero can be interpreted as having a
metro station nearby increase the rate of trip in the afternoon, on
average, when all else is being held constant by a factor of exp(0.009)
= 1.009 .

## Business Implications:

The main takeaways from this model are:

-    Members have a higher rate of trips in the afternoon. Knowing that
    members account for most of the trips, it can explain why there is
    more trip in the afternoon in general.

```{r, echo=FALSE}
df_main %>%    group_by(mem) %>%    summarize(n_tot = sum(n_tot))
```

-   There will be an increase demand on the system in the afternoon
    during the weekend and an increase demand on the system in the
    morning during the weekdays. This could reflect the usage of people
    using bixi to commute to work.

-   There is a strong increase in rate of trips in the afternoon during
    the season of spring, this could be seen as an eagerness for bike
    after winter since afternoon trips seems to be more associated with
    leisure than commuting. Another hypothesis would be that during
    spring the mornings are too cold to bike most often.

-   Following the above hypothesis, as temperature increase, there seems
    to be an increase of rate of trip in the morning. Keep in mind that
    this relation is only true for a given season.

-   Finally concerning the general flow of trips, there seems to be a
    higher rate of departure from stations North West to Parc Lafontaine
    in the morning than in the afternoon. Similarly, we have the inverse
    relation for station in the South East. This means that from an
    operational standpoint, there might be some displacement of bikes
    required from stations to stations depending on the moment of the
    day to keep a balanced fleet of bikes all over the system.

# Research Question 4: Are there significant differences in bike trips counts between weekdays and weekends?

**Objective of Analysis:** The objective of the analysis is to quantify
and assess the differences in ridership during weekend and weekdays. The
analysis would generate insights into the patterns of ridership that
could be useful for operational planning, resource allocation, or
service improvements.

## Variables Selection

To evaluate how time of the day influences trip count, we need to create
a model that accounts for the day being a weekend or a weekday. The
would check if the total trip count as a possion distribution using
chisquare test.

```{r}
chisq.test(table(df_main$n_tot))
```

We can also see from the histogram that the response variable is skewed
to the right. It's clear that linear regression would not be appropriate
in this context. We can fit the Poisson regression model using the glm
function, specifying that the distribution is poisson. We'll start by
fitting the model which includes all explanatory variable:

```{r}
hist(df_main$n_tot, breaks = 20, main = "Histogram of n_tot", xlab = "n_tot")

```

## Model

```{r model_4, echo=FALSE}

# Create a new variable to denote weekdays and weekends
df_main <- df_main %>%
  mutate(weekday_weekend = ifelse(wday %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

# Fit separate GLMs for trip counts by weekday/weekend
total_model <- glm(n_tot ~ weekday_weekend, data = df_main, family = poisson)

# View summaries of the models
summary(total_model)

```

## Interpretation

**Overall Model** R squared, F-stat

**Intercept** :

The intercept (3.004083) represents the log count of the total number of
trips on weekdays.In other words, the number of trips on weekdays is
exp(3.004083)=20.

**Weekend** :

weekend = -0.040604: number of trips is higher on weekdays than weekend.
More specifically, the mean number of trips on weekend is is
exp(-0.040604) = 0.96 times those on weekdays. In other words, the mean
number of trips on weekend is 4% lower than that in weekdays.The p-value
is 3.37e-16 which is less that any acceptable value of alpha which
represent that the difference in trip number during weekend and weekdays
is statistically different.

## Business Implications:

**Resource Allocation**: Given the higher demand during weekdays,
businesses or services dependent on these trips may need to allocate
more resources, such as bikes, maintenance, or staff, during these
periods.

**Pricing Strategies**: If the service is fee-based, adjusting pricing
strategies to account for the difference in demand between weekdays and
weekends could be considered. For instance, offering promotions or
discounts during lower demand days (weekends) to attract more users.

**Marketing Efforts**: Tailoring marketing campaigns or efforts to
encourage more weekend usage could be explored. Promoting special
events, family packages, or leisure-oriented offers during weekends
might help in increasing weekend ridership.

**Operational Optimization**: During weekends, operational adjustments
could be made to enhance the user experience. For instance, ensuring
bike availability, adjusting operating hours, or implementing
user-friendly initiatives to attract more weekend users.

**Service Enhancements**: Understanding the differences in usage
patterns can guide service improvements. Addressing any barriers that
might discourage weekend ridership, such as safety concerns, parking
availability, or service accessibility, could be a focus for
enhancements.

# Limitations and shortcomings

Autocorrelation of data, the observations are not independant, as seens
in our previous analysis.

# Conclusion

# Contribution

Charles Julien :

Gabriel Jobert :

Chike Odenigbo :

Atul Sharma :
