---
title: "Part 4: Linear Mixed Models"
author: "Charles Julien, Chike Odenigbo, Atul Sharma, Gabriel Jobert"
date: "10/20/2023"
geometry: "left=2cm,right=2cm,top=1cm,bottom=1.2cm"
output:
  pdf_document:
    toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, warning=FALSE, include = FALSE, echo=FALSE}
# Import (The fewer the better)

library(tidyverse)
library(ggplot2)
library(dplyr)
library(leaflet)
library(gridExtra)
library(broom)
library(stats)
library(car)
library(janitor)
library(tibble)
```

```{r load_data, include=FALSE}
# Data loading

# Set working directory to file location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
df_raw = read.csv("./../src/bixi10.csv")
df_station = read.csv("./../src/2021_stations.csv")
df_main = left_join(df_raw,df_station,by = c("station"="pk"))
head(df_main)

```

```{r echo=FALSE}
# Data preparation and feature engineering


# Note: Correlation between rev and dur is very high 0.994467

df_main <- df_main %>% mutate(
              holiday = factor(holiday),
              mem = factor(mem),
              mm = factor(mm),
              station = factor(station),
              
              wday_ordered = factor(wday, levels = c("Monday", "Tuesday", "Wednesday",  "Thursday", "Friday", "Saturday", "Sunday"),ordered = TRUE),
              season = ifelse((mm %in% c(12,1,2)), 'Winter',
                       ifelse((mm %in% c(3,4,5)), 'Spring',
                       ifelse((mm %in% c(6,7,8)), 'Summer',
                       ifelse((mm %in% c(9,10,11)), 'Fall','No Season')))),
              
              rain_ind = factor(ifelse((rain != 0), 'Rain', 'NoRain')),
              
              # Compared to Parc Lafontaine 
              North_South = factor(ifelse(latitude>45.5271, 'North', 'South')),
              West_East = factor(ifelse(longitude>-73.5705, 'East', 'West')),
              
              # Beginning of month versus end of month
              PartOfMonth = factor(ifelse(dd>15, 'EOM', 'BOM')),
              
              Metro_ind = factor(ifelse(grepl("Métro", name),1,0)),
              
              lntot = log(n_tot),
              
              rev_per_min = rev/dur,
              rev_per_trip = rev/n_tot,
             
              percent_AM = n_AM/n_tot,
              percent_PM = n_PM/n_tot,
              
              percent_AM_PM_delta = percent_AM - percent_PM,
              n_AM_PM_delta = n_AM - n_PM,
              
              am_pm_ind = ifelse(n_AM>=n_PM, 'AM','PM'),
              
              wknd_ind = factor(ifelse((wday %in% c("Sunday","Saturday")), 'Weekend'
                        ,'Weekday')),
                
              long_wknd_ind = factor(ifelse((wday %in% c("Sunday","Saturday")), 'Weekend'
                        ,ifelse((wday %in% c("Friday","Monday") & holiday=='1'),'Long Weekend','Weekday'))),
        
              year = 2021 ,
              
              date = as.Date(paste(year,mm,dd,sep="/")),
              
              week_num = as.numeric(strftime(date, format = "%V")))
 
```

# Introduction

**Unlocking the Wheels of Urban Mobility: A Data-Driven Analysis of
BIXI**

In the fast-paced, ever-evolving landscape of urban transportation, the
quest to create efficient and sustainable solutions for city dwellers
continues to be a paramount concern. Amidst the diverse array of options
that have emerged in recent years, the BIXI public cycling service
stands as a beacon of sustainable urban mobility. Offering an
accessible, convenient, and eco-friendly mode of transportation, BIXI
has transformed the way people navigate and experience cities.

As part of our commitment to understanding and improving urban
transportation systems, our consultant team has embarked on an in-depth
exploration of BIXI's operational data. The objective of this report is
to provide a comprehensive analysis of the data collected from the BIXI
service. By leveraging statistical and data analysis techniques, we aim
to uncover valuable insights into the usage patterns, financial
dynamics, and various factors affecting BIXI's performance. Our study
covers an extensive range of factors, including ridership trends,
environmental conditions, user classifications, and more.

One of the central questions addressed in this report is whether revenue
generated by the BIXI service and trip duration significantly varies
during weekends compared to weekdays. we also delve into the other
factors affecting the duration of trips and the revenue generated by
non-members. Our methodology combines data analysis, data visualization,
and statistical modeling, with a primary focus on using R, a powerful
statistical tool, to extract meaningful information from the BIXI
dataset.

By analyzing this data, we aim to assist BIXI in making data-informed
decisions to enhance the efficiency and quality of their services,
ultimately contributing to the betterment of urban living. We believe
that the findings and recommendations presented in this report will not
only provide valuable insights to BIXI but also serve as a valuable
reference for urban planners, researchers, and policymakers who are
dedicated to creating more sustainable, convenient, and enjoyable urban
environments.

The following sections of the report will delve into the specifics of
our data analysis, share our findings, and provide recommendations based
on the insights gathered during this project.

# Business/Research questions

-   Research Question 1: How do seasonal factors impact trip revenue for
    BIXI Montréal?
-   Research Question 2: How do daily and weekly patterns impact trip
    durations for BIXI Montréal?
-   Research Question 3: What variables impact the average bixi trip
    duration?

# Pre-processing

## Imputation

Find a better imputation method

# Research Question 1: How do seasonal factors impact trip revenue for BIXI Montréal?
ADD INTERACTION TERM
**Objective of Analysis:** This regression model is examining the impact
of the month (`mm`), average daily temperature (`temp`), and total
amount of rainfall (`rain`) and membershipt (`mem`) on the revenue
(`rev_pred`) generated by trips leaving from a specified station.

## Model

```{r Seasonal_effect_revenue, echo=FALSE}
seasonal_effect_rev_model <- lm(rev_pred ~ mm + temp + rain + mem, data = df_main)
summary(seasonal_effect_rev_model)
```

## Interpretation

**Intercept**: The intercept is -13.9996. This means that when all other
predictors are zero (for non member on april), the predicted revenue is
-\$13.99. However, this should be treated with caution since an
intercept in this context doesn't have a clear business interpretation.

**Seasonality (Month):** - The revenue seems to have a seasonal pattern.
Compared to April (reference month), the model suggests that there's an
increase in predicted revenue from May (mm5) through September (mm9).
The highest increase in predicted revenue relative to April is in
September (mm9) with an increase of \$41.58 on average. After September,
there is a decline in October (mm10) and November (mm11), suggesting
that as we move into cooler months, the revenue decreases, but it's
still higher than in April.

**Temperature (`temp`):** For every degree Celsius increase in
temperature, the predicted revenue increases by approximately \$1.02 on
average. This suggests that warmer days tend to generate more revenue.

**Rainfall (`rain`):** - For every additional mm of rainfall, the
revenue decreases by approximately \$1.26 on average. This is intuitive
as rainy days likely deter users from renting bikes, leading to reduced
revenue.

**Membership**: If a rider is a member (mem1), the predicted revenue
increases by \$72.88 on average compared to non-member. This means that
members contribute significantly more to the revenue compared to
non-members.

**Model Fit:** - The model has an R-squared value of 0.2782, indicating
that approximately 27.82% of the variation in predicted revenue is
explained by the model. - The F-statistic and its associated p-value
(which is very close to zero) suggest that the model is statistically
significant, and at least one of the predictors is useful in predicting
revenue.

## Business implications

1.  **Operational Adjustments:** Given that revenue is higher in warmer
    months, consider optimizing operations for this period. This might
    involve higher staffing, more promotional activities, or ensuring
    optimal equipment availability.

2.  **Rainy Day Strategies:** Since rainfall seems to negatively impact
    revenue, consider implementing strategies to mitigate this. For
    instance, promotional offers or special activities/events for rainy
    days might help attract customers.

3.  **Membership** : While temperature and rain have expected impacts on
    bike rentals, it's notable how significant the role of membership is
    in driving revenue. Bixi should consider incentivizing memberships,
    given the clear revenue increase associated with members.

## Verification of Assumptions
NEED A MORE COMPLETE ASSUPTION VALIDATION
### Normality of residuals

```{r echo=FALSE}
par(mfrow=c(1,2), pin=c(2,2))  # 1 row, 2 columns
seasonal_effect_rev_residual_qqplot = qqnorm(seasonal_effect_rev_model$residuals)
qqline(seasonal_effect_rev_model$residuals)
seasonal_effect_rev_residual_hist = hist(seasonal_effect_rev_model$residuals, main="Histogram of Residuals", xlab="Residuals")

```

1.  **Histogram of Residuals**: These histograms have a clear right-skew
    with a peak close to zero and a long tail towards the right. This
    suggests that most residuals are clustered around zero, but there
    are a few larger positive residuals. This is an indication that the
    normality assumption of the residuals may be violated.

2.  **Normal Q-Q Plot**: Most of the points are close to the line, which
    is a good sign. However, there's a clear deviation from the line on
    the top right corner, suggesting the presence of larger residuals
    that are not explained by a normal distribution. This reiterates the
    presence of the right skew seen in these histograms.

**Overall Interpretation**: The residuals are not perfectly normal. They
show a positive skewness, indicating there might be some observations
with higher residuals (perhaps outliers or instances where the model
systematically underpredicts). The deviation from normality might not be
a problem because the sample is large enough.

# Research Question 2: How do daily and weekly patterns impact trip durations for BIXI Montréal?

**Objective of Analysis:** This regression model is examining the impact
of the day of the month (`dd`), day of the week (`wday`), and holidays
(`holiday`) on the revenue (`rev`) generated by trips leaving from a
specified station.

## Model
DONT USE DD AS CONTINUOUS
```{r Time_pattern_trip_duration, echo=FALSE}
time_pattern_dur_model <- lm(dur ~ dd + wday + holiday, data=df_main)
summary(time_pattern_dur_model)
```

## Interpretation

**Overall Model** : The model explains about 0,5% of the variability in
total rental durations. The F-statistic and its associated p-value
confirm that the model is statistically significant and that at least
some of the predictors have significant effects.

**Intercept (284.98)***: - On an average day (specifically, a
non-holiday), the expected rental duration is approximately 4h45. - This
value is statistically significant (*\*\* p-value \< 2e-16), which
indicates strong evidence against the null hypothesis.

**Day of the Month (`dd`)**: - For each additional day in the month, the
rental duration increases by an average of 0.04 minutes. - This effect
is not statistically significant (p-value = 0.9144), suggesting the day
of the month might not be a meaningful predictor for the duration of
BIXI bike rentals.

**Day of the Week (`wday`)**: - Compared to Fridays: - Rentals on
**Mondays** are, on average, **49 minutes shorter**. This is
statistically significant (\*\*\* p-value = 2.55e-05). - Rentals on
**Saturdays** are about **15.05 minutes longer** on average, but this is
not statistically significant (p-value = 0.1868). - Rentals on
**Sundays** are about **6.51 minutes shorter** on average, but this is
also not statistically significant (p-value = 0.5711). - Rentals on
**Thursdays** are **16.95 minutes shorter** on average, but this isn't
statistically significant either (p-value = 0.1428). - Rentals on
**Tuesdays** are **34.48 minutes shorter** on average, and this is
statistically significant (\*\* p-value = 0.0030). - Rentals on
**Wednesdays** are **20.57 minutes shorter** on average. This result is
at the borderline of significance (p-value = 0.0745).
DO A GLOBAL F TEST
**Holiday (`holiday1`)**: - On holidays, bike rentals are, on average,
**1 hour and 8 minutes longer** compared to non-holidays. This is
statistically significant (\*\* p-value = 0.0012), suggesting that
holidays have a meaningful impact on the duration of bike rentals.

**Business Implications**: BIXI bike rentals tend to be shorter on
Mondays and Tuesdays compared to Fridays, and rentals on holidays are
significantly longer than on non-holidays. Planning for resource
allocation, marketing strategies, or promotional campaigns should
consider these patterns to optimize business operations. Keeping in mind
the higher trip duration during holidays, adequate steps should be taken
to ensure Bixi availability at stations.

## Verification of assumptions
NOT COMPLETE
### Normality of residuals

```{r echo=FALSE}
par(mfrow=c(1,2), pin=c(2,2))  # 1 row, 2 columns

time_pattern_dur_residual_qqplot = qqnorm(time_pattern_dur_model$residuals)
qqline(time_pattern_dur_model$residuals)
time_pattern_dur_residual_hist = hist(time_pattern_dur_model$residuals, main="Histogram of Residuals", xlab="Residuals")

```

**Normal Q-Q Plot**: From the given Q-Q plot, the points deviate
significantly from the diagonal line, especially in the tails. This
suggests that the residuals are not normally distributed. The heavy
tails (points deviating from the line at both ends) suggest the presence
of potential outliers or extreme values in the residuals.

**Histogram of Residuals**: The histogram shows that the majority of
residuals are clustered around zero, but there are some extreme positive
values. This is consistent with the observation from the Q-Q plot and
indicates a possible right-skewed distribution of residuals. In the
context of regression, the CLT means that even if the residuals aren't
perfectly normally distributed in the population, the sampling
distribution of the regression coefficients will be approximately normal
if the sample size is large enough.

**Implications**:

-   While large sample sizes can make the assumption of normality less
    crucial, it doesn't mean analysts should ignore violations of other
    assumptions or entirely disregard the distribution of residuals.
    Diagnostics and plots (like Q-Q plots) still provide valuable
    information about potential model misspecifications or the presence
    of influential outliers.

-   Moreover, a large sample size can sometimes detect statistically
    significant relationships even when they are practically
    insignificant. So, while p-values might be small, the effect sizes
    or coefficients might not be practically meaningful.

# Research Question 3: What variables impact the average bixi trip duration?

**The objective** is to identify the driving factors of a bixi's trip
length when we control for most of the variables. Trip length is one of
the three important variables that drives revenue, the other ones being
the number of trips and the pricing scheme. Keep in mind that increasing
the trip length does not necessarily increase revenues since an unwanted
increase in trip length may discourage users from using bixi's system
and result in a decrease in trip number.

## Variables Selection
REMOVE PART OF MONTH
Our goal is to incorporate most of the important variables in order to
increase our chance of respecting the assumption of E(e)=0 and thus
making our model more telling.

Variables that make business sense to include:

From our seasonality analysis we identified:

-   Season; grouping of months from april to november in their
    respective season (`season`)

-   Temperature in degrees celcius (`temp`)

-   Rainfall in mm (`rain`)

From our daily and weekly pattern analysis we identified:

-   Part of the week i.e. weekend or weekday (`wknd_ind`)

-   If it is a holiday (`holiday`)

Some other variables that are interesting:

-   If the user is a member(`mem`)

-   Location of the bixi station compared to Parc Lafontaine, a landmark
    in the middle of the bixi station system (`North_South`) and
    (`West_East`)

-   Proportion of trips in the morning versus the whole day
    (`percent_AM`)

-   If the station name contains the word 'metro' (`Metro_ind`)

-   If we are at the beginning of month(before the 15th) or the end of
    the month (`PartOfMonth`)

**Interactions**: In our EDA we observed a different week day usage of
the member and non members, thus an interaction term between members and
day of week would be interesting. (`wday*mem`).

**Correlation:**

Let's take a quick look at the correlation between our numerical
variables to estimate the effect of collinearity.

```{r , echo=FALSE}
cor(df_main[c('avg', 'temp', 'rain', 'n_tot', 'percent_AM')])
```

We see very low correlation between the Xs which means we should not get
any problems with collinearity between our numerical variables.

After the assumptions verification we chose to exclude `n_tot` :
VERIFY
```{r , echo=FALSE, fig.width=4, fig.height=3}
plot(df_main$n_tot, df_main$avg)  
```

The variable total number of trip (`n_tot`) has been removed from the
regression because it did not pass the assumption of constant variance,
making the model not correctly specified. This makes intuitive sense
since as the number of trip increases, the average trip duration should
converge towards the true mean.

## Model
INTERACTIO AND QUADRANT
```{r , echo=FALSE}
model_AvgDur <- lm(avg ~ season + temp + rain + wknd_ind*mem + holiday  + North_South + West_East + percent_AM + PartOfMonth + Metro_ind, data = df_main)
summary(model_AvgDur)
```

## Interpretation

**Overall Model** - The model explains approximately 10% of the
variation in the average trip duration which means that other factors
are also at play and are not included in the model. The p value
associated with the F-statistic is very low, hence our model is globally
significant.

**Intercept** : The interpretation of the intercept does not make sense
in this case 

**Season**: The reference level is fall. We can see that on average trip
duration during spring and summer are respectively 2.7 and 0.5 minutes
longer than in fall holding everything else constant.

**Temperature**: The coefficient of temperature is 0.11 which means that
an increase in temperature of 1 degree celcius corresponds to an
increase of average trip duration of 0.11 minutes on average holding all
else constant.

**Rainfall**: The coefficient for rain is -0.1 which means that an
increase in rainfall of 1 mm corresponds to a decrease of average trip
duration of 0.1 minutes on average holding all else constant.

**Effect of Weekend Indicator and membership**: Since there exists an
interaction between both variables, it is no longer possible to
interpret one without the other. This implies that the relation between
average trip duration and membership is different depending on the
moment of the week. The opposite is also true, the relation between
average trip duration and the moment of the week is different depending
on the membership status. We observe that non-member have longer trips
on average and that weekend trips tends to increase average trip length.

The 4 different levels in order of trip length are as follows:
EXPLAIN IN WORDS
1.  Lowest level : Weekday and member (-1.84 minutes)

2.  Second lowest : Weekend and member (2.5-1.8-1.5 = -0.8 minutes)

3.  Reference level: Weekday and non-member (0 minutes)

4.  Highest level : weekend and non-member (+2.47 minutes)

**Holiday**: The coefficient for holiday is 1.06 which means that during
holidays average trip duration is 1.06 minutes higher on average than
during non-holidays, holding all else constant.

**North_South and West_East**: Their coefficients are 0.08 and -0.26
which means that on average the average trip duration for trips starting
at a station South of Parc Lafontaine or West is 0.08 and -0.26 minutes
different from their counter parts respectively, holding all else
constant. Keep in mind that the coefficient for North_South is not
significantly different from zero

**Percent AM**: The magnitude of the coefficient -2.03 is less important
than its sign for our interpretation. What it means is that as the
proportion of trips in the morning increases, the average trip duration
generally decreases when holding all else constant. This hints that
trips in the morning might be shorter on average than trip in the
afternoon, hence bring in less revenue.

**Part of Month** : The coefficient for part of month is -0.22 which
means that on average, the average trip duration is 0.22 minutes shorter
in the second half of the month holding all else constant. This feature
was not found to be significantly different from zero.

**Metro Indicator** : Metro indicator's coefficient is -0.75 which means
that the expected value for average trip length decreases by 0.75
minutes when a bixi station is near a metro acces point, holding all
else constant. This would suggest that user who rent bikes after making
a metro ride are closer to their final destination than in other cases.

## Business Implications:

1.  **Promotion and Marketing**: For the same temperature, average trip
    length tends to be the longest in spring. This indicates that users
    are eager to use bikes after winter. This insight could be used for
    promotion purposes.

2.  **Resource Allocation**: Expect longer trips when it is hot and
    non-rainy outside. Even more if it is a weekend or holiday. Also,
    bikes tend to be borrowed longer during the afternoon than in the
    morning. Stations south of Parc Lafontaine have on average longer
    trip duration, which may suggests that stations are further from one
    another. There might be some space for additional stations.

3.  **Pricing Strategy**: The usage that is associated with the longest
    trip length based on our interaction term is for non-members during
    the weekend. Charging a heftier price for these people at that time
    may increase profit margins significantly.

## Verification of assumptions and collinearity

### Variance Inflation Factor

Let's use the variance inflation factor to verify for collinearity, we
will use a standard threshold of 5.

```{r , echo=FALSE}
vif(model_AvgDur)
```

No major problem is detected, since the global vifs are all relatively
low.

### Verification of Normality of Residuals

```{r, echo=FALSE, fig.width=4, fig.height=3}
#pin=c(2,2)
res = hist(model_AvgDur$residuals, main="Histogram of Residuals", xlab="Residuals")

```

No problem here, residuals are normally distributed.

### Model correctly specified

```{r , echo=FALSE, fig.width=4.5, fig.height=3.5}
plot(predict(model_AvgDur), resid(model_AvgDur)) 
```

The model seems to be correctly specified.

### Verificaiton of Heteroscedasticity

```{r , echo=FALSE}
#pin=c(2,2)
residuals_df <- data.frame(df_main, residuals = residuals(model_AvgDur))

p1 <- ggplot(data = residuals_df, aes(x = temp, y = residuals)) +
  geom_point() +
  labs(x = "temp", y = "Residuals")

p2 <- ggplot(data = residuals_df, aes(x = rain, y = residuals)) +
  geom_point() +
  labs(x = "rain", y = "Residuals")

p3 <- ggplot(data = residuals_df, aes(x = n_tot, y = residuals)) +
  geom_point() +
  labs(x = "n_tot", y = "Residuals")

p4 <- ggplot(data = residuals_df, aes(x = percent_AM, y = residuals)) +
  geom_point() +
  labs(x = "percent_AM", y = "Residuals")

p5 <- ggplot(data = residuals_df, aes(x = season, y = residuals)) + geom_boxplot() +
  labs(x = "season", y = "Residuals")

p6 <- ggplot(data = residuals_df, aes(x = mem, y = residuals)) + geom_boxplot() +
  labs(x = "mem", y = "Residuals")

p7 <- ggplot(data = residuals_df, aes(x = holiday, y = residuals)) + geom_boxplot() +
  labs(x = "holiday", y = "Residuals")

p8 <- ggplot(data = residuals_df, aes(x = North_South, y = residuals)) + geom_boxplot() +
  labs(x = "North_South", y = "Residuals")

p9 <- ggplot(data = residuals_df, aes(x = West_East, y = residuals)) + geom_boxplot() +
  labs(x = "West_East", y = "Residuals")

p10 <- ggplot(data = residuals_df, aes(x = PartOfMonth, y = residuals)) + geom_boxplot() + labs(x = "PartOfMonth", y = "Residuals")

p11 <- ggplot(data = residuals_df, aes(x = Metro_ind, y = residuals)) + geom_boxplot() + labs(x = "Metro_ind", y = "Residuals")

grid.arrange(p1, p2, p4, p5, p6, p7, p8, p9, p10, p11, ncol=3)

```

No major problem of heteroscedasticity were detected. The variable
`n_tot` has been removed as stated earlier.


# Limitations and shortcomings

-   Causation vs. Correlation: The regression model captures
    relationships but does not establish causation.
-   Data Exclusions: The data only considers trips under 60 minutes,
    which might exclude a segment of users who use BIXI for longer
    journeys.
-   Other External Factors: Events, road conditions, or public
    transportation disruptions can affect BIXI usage but are not
    captured in the dataset.

# Conclusion

In conclusion, several key operational and strategic considerations have
emerged from the data analysis of BIXI bike rentals:

Operational Adjustments: The data suggests that revenue is higher in
warmer months. To capitalize on this, it is advisable to optimize
operations during this period, which could involve increasing staffing,
enhancing promotional activities, and ensuring optimal equipment
availability.

Rainy Day Strategies: Rainfall appears to have a negative impact on
revenue. Implementing strategies to mitigate this effect, such as
promotional offers or special activities for rainy days, may help
attract more customers.

Promotion and Marketing: Data indicates that average trip length is
longest in spring, suggesting an eagerness to use bikes after winter.
This insight can be leveraged for promotional purposes.

Resource Allocation: Understanding patterns in trip duration based on
weather, time of day, and location is crucial for resource allocation.
Longer trips are expected during hot, non-rainy weekends and holidays.
Stations in certain areas have longer trip durations, indicating
potential for additional station placement.

Pricing Strategy: The analysis highlights that non-members on weekends
tend to take longer trips. Adjusting pricing for this group during these
times could significantly increase profit margins.

Operational Strategy: It's important to consider the tradeoff between
the number of trips and average trip length. Increasing the number of
trips on a given day may lead to shorter hauls. This information should
inform operational decisions. Incorporating these insights into the
business's operations and strategies can lead to improved efficiency,
customer satisfaction, and profitability

# Contribution

Charles Julien :Research question 3, version control, part of feature
engineering, formating.

Gabriel Jobert : Research question 1 and 2

Chike Odenigbo: exploratory models (not included), feature engineering,
influential observations, autocorrelation

Atul Sharma: Contributed in developing the Reseacrh questions,
interpreting the findings of the model and finalising the conclusion .
